---
title: "Time Series Forecasts"
author: "Aleksei Prishchepo"
date: "2025-11-16"
fig-format: svg
number-sections: true
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    highlight-style: tango
---

In this notebook we're going to create forecasts for the time series
data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.width = 8)
```

## Load Data

In Power BI we will reference the Events table. Here we will load the
data from a CSV file.

```{r load-data}
library(readr)
library(dplyr)

dataset <- read_csv("Events.csv")
dataset <- dataset |>
  mutate(Date = format(as.Date(Date), "%d.%m.%Y"))
dataset |> glimpse()
```

## Forecast by Country

```{r forecast-by-country}

library(dplyr)
library(lubridate)
library(forecast)

dataset <- dataset |> rename(ds = Date)
dataset$ds <- as.Date(dataset$ds, format = "%d.%m.%Y")

earliest_date <- min(dataset$ds)
latest_date <- max(dataset$ds)

end_of_year <- as.Date(paste0(year(latest_date), "-12-31"))

# Forecast horizon in days
h_days <- difftime(end_of_year, latest_date, units = "days") |> as.numeric()

# Aggregate by ds and country
daily <- dataset |>
  group_by(country, ds) |>
  summarise(revenue = sum(net_revenue_usd), .groups = "drop")

# Define modeling function for each country
forecast_country <- function(df) {
  country_name <- unique(df$country)

  # Fill missing dates for this country
  all_days <- tibble(ds = seq(earliest_date, latest_date, by = "day"))
  df_full <- all_days |>
    left_join(df, by = "ds") |>
    mutate(
      country = country_name,
      revenue = ifelse(is.na(revenue), 0, revenue)
    )

  # Skip if too few data points
  if (nrow(df_full) < 100) {
    return(
      tibble(
        country = country_name,
        ds = df_full$ds,
        actual = df_full$revenue,
        forecast = NA_real_,
        lower95 = NA_real_,
        upper95 = NA_real_
      )
    )
  }

  # Convert to daily time series
  ts_data <- ts(df_full$revenue, frequency = 7) # weekly cycle assumed

  # Fit model
  model <- auto.arima(ts_data)
  # model <- tbats(ts_data)

  fc <- forecast(model, h = h_days)

  # Combine historical and forecasted
  result <- tibble(
    country = country_name,
    ds = seq(earliest_date, by = "day", length.out = length(ts_data) + h_days),
    actual = c(df_full$revenue, rep(NA, h_days)),
    forecast = c(rep(NA, length(ts_data)), as.numeric(fc$mean)),
    lower95 = c(rep(NA, length(ts_data)), fc$lower[, 2]),
    upper95 = c(rep(NA, length(ts_data)), fc$upper[, 2])
  )
  return(result)
}

output <- daily |>
  group_by(country) |>
  group_modify(~ forecast_country(.x)) |>
  ungroup()

# Add rolling mean for smoother actual trend line
output <- output |>
  group_by(country) |>
  arrange(ds) |>
  mutate(
    actual_ma = zoo::rollmean(actual, k = 90, fill = NA, align = "right")
  ) |>
  ungroup()

output <- output |> rename(Date = ds)

output |> tail()
```

Plot ACF/PACF

```{r}
library(forecast)
tsdisplay(ts(daily |> filter(country == "Australia") |> pull(revenue), frequency = 7), main = "ACF for Australia Revenue")
```

Plot the results for a sample country.

```{r fig-plot-country-forecast, fig.height=5, fig.width=10}
library(ggplot2)

sample_country <- "Spain"
df_plot <- output |> filter(country == sample_country)
ggplot(df_plot, aes(x = Date)) +
  geom_line(aes(y = actual), color = "blue") +
  geom_line(aes(y = forecast), color = "red") +
  geom_line(aes(y = actual_ma), color = "green", linetype = "dashed") +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), alpha = 0.2, fill = "orange") +
  labs(
    title = paste("Revenue Forecast for", sample_country),
    y = "Revenue (USD)",
    x = "Date"
  ) +
  theme_minimal()


```

## Forecast by Channel

here we repeat the same process but group by sales channel.

```{r load-data-channel}
dataset <- read_csv("Events.csv")
dataset <- dataset |>
  mutate(Date = format(as.Date(Date), "%d.%m.%Y"))
```

```{r forecast-by-channel}

library(dplyr)
library(lubridate)
library(forecast)

dataset <- dataset |> rename(ds = Date)
dataset$ds <- as.Date(dataset$ds, format = "%d.%m.%Y")

earliest_date <- min(dataset$ds)
latest_date <- max(dataset$ds)

end_of_year <- as.Date(paste0(year(latest_date), "-12-31"))

# Forecast horizon in days
h_days <- difftime(end_of_year, latest_date, units = "days") |> as.numeric()

# Aggregate by ds and channel
daily <- dataset |>
  group_by(channel, ds) |>
  summarise(revenue = sum(net_revenue_usd), .groups = "drop")

# Define modeling function for each country
forecast_channel <- function(df) {
  channel_name <- unique(df$channel)

  # Fill missing dates for this country
  all_days <- tibble(ds = seq(earliest_date, latest_date, by = "day"))
  df_full <- all_days |>
    left_join(df, by = "ds") |>
    mutate(
      channel = channel_name,
      revenue = ifelse(is.na(revenue), 0, revenue)
    )

  # Skip if too few data points
  if (nrow(df_full) < 100) {
    return(
      tibble(
        channel = channel_name,
        ds = df_full$ds,
        actual = df_full$revenue,
        forecast = NA_real_,
        lower95 = NA_real_,
        upper95 = NA_real_
      )
    )
  }

  # Convert to daily time series
  ts_data <- ts(df_full$revenue, frequency = 7) # weekly cycle assumed

  # Fit model
  model <- auto.arima(ts_data)
  # model <- tbats(ts_data)

  fc <- forecast(model, h = h_days)

  # Combine historical and forecasted
  result <- tibble(
    channel = channel_name,
    ds = seq(earliest_date, by = "day", length.out = length(ts_data) + h_days),
    actual = c(df_full$revenue, rep(NA, h_days)),
    forecast = c(rep(NA, length(ts_data)), as.numeric(fc$mean)),
    lower95 = c(rep(NA, length(ts_data)), fc$lower[, 2]),
    upper95 = c(rep(NA, length(ts_data)), fc$upper[, 2])
  )
  return(result)
}

output <- daily |>
  group_by(channel) |>
  group_modify(~ forecast_channel(.x)) |>
  ungroup()

# Add rolling mean for smoother actual trend line
output <- output |>
  group_by(channel) |>
  arrange(ds) |>
  mutate(
    actual_ma = zoo::rollmean(actual, k = 90, fill = NA, align = "right")
  ) |>
  ungroup()

output <- output |> rename(Date = ds)

output |> tail()
```

Plot the results for a sample channel

```{r fig-plot-channel-forecast, fig.height=5, fig.width=10}
library(ggplot2)

sample_channel <- "Direct Sales"
df_plot <- output |> filter(channel == sample_channel)
ggplot(df_plot, aes(x = Date)) +
  geom_line(aes(y = actual), color = "blue") +
  geom_line(aes(y = forecast), color = "red") +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), alpha = 0.2, fill = "orange") +
  labs(
    title = paste("Revenue Forecast for", sample_channel),
    y = "Revenue (USD)",
    x = "Date"
  ) +
  theme_minimal()


```

## Forecast by Category

Next, we will forecast by product category. For this we need to load the
product data to get the category information.

```{r load-data-category}
dataset <- read_csv("Events.csv")
dataset <- dataset |>
  mutate(Date = format(as.Date(Date), "%d.%m.%Y"))

file_name <- "DataDNA Dataset Challenge - E-commerce Dataset - November 2025.xlsx"
products <- readxl::read_xlsx(file_name, sheet = 2) |>
  select(product_id, category)

dataset <- dataset |> left_join(products, by = "product_id")

```

The code below will be added as a step in Power Query.

```{r forecast-by-category}

library(dplyr)
library(lubridate)
library(forecast)

dataset <- dataset |> rename(ds = Date)
dataset$ds <- as.Date(dataset$ds, format = "%d.%m.%Y")

earliest_date <- min(dataset$ds)
latest_date <- max(dataset$ds)

end_of_year <- as.Date(paste0(year(latest_date), "-12-31"))

# Forecast horizon in days
h_days <- difftime(end_of_year, latest_date, units = "days") |> as.numeric()

# Aggregate by ds and category
daily <- dataset |>
  group_by(category, ds) |>
  summarise(revenue = sum(net_revenue_usd), .groups = "drop")

# Define modeling function for each category
forecast_category <- function(df) {
  category_name <- unique(df$category)

  # Fill missing dates for this category
  all_days <- tibble(ds = seq(earliest_date, latest_date, by = "day"))
  df_full <- all_days |>
    left_join(df, by = "ds") |>
    mutate(
      category = category_name,
      revenue = ifelse(is.na(revenue), 0, revenue)
    )

  # Skip if too few data points
  if (nrow(df_full) < 100) {
    return(
      tibble(
        category = category_name,
        ds = df_full$ds,
        actual = df_full$revenue,
        forecast = NA_real_,
        lower95 = NA_real_,
        upper95 = NA_real_
      )
    )
  }

  # Convert to daily time series
  ts_data <- ts(df_full$revenue, frequency = 7) # weekly cycle assumed

  # Fit model
  model <- auto.arima(ts_data)
  # model <- tbats(ts_data)

  fc <- forecast(model, h = h_days)

  # Combine historical and forecasted
  result <- tibble(
    category = category_name,
    ds = seq(earliest_date, by = "day", length.out = length(ts_data) + h_days),
    actual = c(df_full$revenue, rep(NA, h_days)),
    forecast = c(rep(NA, length(ts_data)), as.numeric(fc$mean)),
    lower95 = c(rep(NA, length(ts_data)), fc$lower[, 2]),
    upper95 = c(rep(NA, length(ts_data)), fc$upper[, 2])
  )
  return(result)
}

output <- daily |>
  group_by(category) |>
  group_modify(~ forecast_category(.x)) |>
  ungroup()

# Add rolling mean for smoother actual trend line
output <- output |>
  group_by(category) |>
  arrange(ds) |>
  mutate(
    actual_ma = zoo::rollmean(actual, k = 90, fill = NA, align = "right")
  ) |>
  ungroup()

output <- output |> rename(Date = ds)

output |> tail()
```

Plot the results for a sample category

```{r fig-plot-category-forecast, fig.height=5, fig.width=10}
library(ggplot2)

sample_category <- "AI Productivity"
df_plot <- output |> filter(category == sample_category)
ggplot(df_plot, aes(x = Date)) +
  geom_line(aes(y = actual), color = "blue") +
  geom_line(aes(y = forecast), color = "red") +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), alpha = 0.2, fill = "orange") +
  labs(
    title = paste("Revenue Forecast for", sample_category),
    y = "Revenue (USD)",
    x = "Date"
  ) +
  theme_minimal()

```

## References

-   [Building an E-Commerce Dashboard with Power BI and R –
    Blog](https://blog.frequentist.org/posts/20251126-e-commerce-dashboard/)
-   [Repository on
    GitHub](https://github.com/AxesAccess/DataDNA-Dataset-Challenge-E-commerce-Dataset-November-2025)
    
